Finally Tagless, Partially Evaluated

### 2007-06-12 tagless/tagless-final.tex
17.5 pages.

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.

@@

(b) \begin{code2}
let test1 = A (L (V VZ), B true)
\end{code2}

(c) \begin{code2}
let rec lookup (x::env) = function
| VZ   -> x
| VS v -> lookup env v

let rec eval0 env = function
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code2}

(d) \begin{code2}
type u = UB of bool | UA of (u -> u)
\end{code2}

(e) \begin{code2}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code2}

(f) \begin{code2}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code2}

(g) \begin{code2}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code2}

(h) \begin{code2}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code2}
\caption{OCaml code illustrating the tag problem.}
\label{fig:tag-problem}
\end{figure}
It is straightforward to create an algebraic data type, say in OCaml%
%aplas-specific
Fig.~\ref{fig:tag-problem}(a)%
, to
represent object terms such as those in Figure~\ref{fig:object}.
% fig:tag-problem part (a) was here
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
%aplas-specific
in Fig.~\ref{fig:tag-problem}(b)%
% fig:tag-problem part (b) was here

Following \cite{WalidICFP02},
we try to implement an interpreter function |eval0|%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(c)%
. It takes
an object term such as |test1| above and gives us its value.
The first argument to |eval0| is the environment, initially empty,
which is the list of values bound to free variables in the
interpreted code.
% fig:tag-problem part (c) was here
If our OCaml-like metalanguage were untyped, the code above would be 
acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \citep[\S1.1--2]{WalidICFP02}.

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(d)%
.
% fig:tag-problem part (d) was here
We can then write a typed interpreter%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(e)%
% fig:tag-problem part (e) was here
whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(f)%
.
% fig:tag-problem part (f) was here
The unfortunate tag |UB| in the result reflects that |eval| is a partial
function.  First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(g), 
if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
% fig:tag-problem part (g) was here
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(h)%
.
% fig:tag-problem part (h) was here
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.

@@

\subsection{Solutions using fancier types}

It is commonly thought that to interpret a typed object language in
a typed metalanguage while preserving types is difficult and requires
GADTs or dependent types \citep{taha-tag}.  In fact, this problem
motivated much work on GADTs \cite{xi-guarded,peyton-jones-simple} and
on dependent types \cite{WalidICFP02,fogarty-concoqtion}.

For a metalanguage's type system to allow the well-typed object term
|test1| but disallow the ill-typed object term |test2|, fancier types
such as GADTs or dependent types seem necessary.  Yet other type systems
have been proposed to distinguish closed terms like |test1| from open
terms like |test3|
%aplas \cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01},
so that |lookup| never receives an empty environment.  We discuss these
proposals further in \S\ref{related}; here we just note that many
advanced type systems have been devised to ensure statically that an
object term is well-typed and closed.

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions%
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(a)%
.
% fig:tag-final-proposal part (a) was here
These functions comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
%aplas-specific
Fig.~\ref{fig:tag-final-proposal}(b)%
.
% fig:tag-final-proposal part (b) was here
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment%
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(c)%
.
% fig:tag-final-proposal part (c) was here
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as |testf3| below gives a type error rather than
a run-time error
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(d)%
.
% fig:tag-final-proposal part (d) was here
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.
\begin{figure}
%
(a) \begin{code2}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code2}

(b) \begin{code2}
let testf1 = app (lam varZ) (b true)
\end{code2}

(c) \begin{code2}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code2}

(d) \begin{code2}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code2}
\caption{OCaml code illustrating our final proposal.}
\label{fig:tag-final-proposal}
\end{figure}
%
In sum, by Church\hyp encoding terms using ordinary functions, we
achieve a tagless evaluator for a typed object language in a
metalanguage with a simple 
%aplas
Hindley-Milner 
type system
%aplas \cite{hindley-principal,milner-theory}
.  In this \emph{final} rather
than \emph{initial} approach, both kinds of run-time errors in
\S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter
uses no universal type or pattern matching, it never results in a
run-time error (and is in fact total).  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \cite{WalidICFP02} of error checking.
\Citet{Gluck-jones-optimality} explains deeper technical reasons that
inevitably lead to these performance penalties.

@@

(c) \begin{code2}
testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code2}
\caption{Symantics in Haskell.}
\label{fig:symantics-haskell}
\end{figure}
% fig:symantics-haskell part (a) was here
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For another example, the classical $\mathit{power}$ function is
%aplas-specific
in Fig.~\ref{fig:symantics-haskell}(b)
% fig:symantics-haskell part (b) was here
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
in Fig.~\ref{fig:symantics-haskell}(c).
% fig:symantics-haskell part (c) was here
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag.
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete implementations.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation
\begin{code3}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
: \texttt{module EXR = EX(R)}%
.
%% \begin{code}
%% module EXR = EX(R)
%% \end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
% aplas: commented out
\begin{comment}
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
\end{comment}
%aplas: the following line is for aplas only
It is obvious to the compiler that
%
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
These propositions follow immediately from the soundness of the
metalanguage's type system.

% aplas-specific
For variety, the source code shows another interpreter, |L|, which measures
the \emph{size} of each object term, defined as the number of term
constructors. This interpreter is not only tagless but also total. It
``evaluates'' even seemingly divergent terms.
\begin{comment}
For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the expression
\begin{code}
let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
\begin{code}
app (fix (fun self -> self)) (int 1)
\end{code}

@@

(d) \begin{code2}
.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
   n_4 in self_3) x_1 7>.
\end{code2}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
%
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(b),
% fig:interpreter-C part (b) was here
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
with
%aplas-specific
Fig.~\ref{fig:interpreter-C}(c),
% fig:interpreter-C part (c) was here
gives a code value with many apparent $\beta$-redexes
%aplas-specific
Fig.~\ref{fig:interpreter-C}(d)
% fig:interpreter-C part (d) was here

This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and does not
  call the interpreter, unlike the recursive calls to |eval0| and
  |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.
%aplas: the line is for aplas. The original paragraph commented out
The accompanying code shows the corresponding Haskell implemention. 
\begin{comment}
We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
  (TH). The output of our compiler will then be assuredly type-correct TH.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.
\end{comment}

@@

Roughly, a partial evaluator interprets each object term to either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, one might define
the Haskell data type
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

@@

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}

@@

A lot of effort has gone into ``typing dynamic typing'': to statically
type-check dynamically\hyp typed values
\cite{baars-typing,haskell-list},
using the host language's type system to varying extents.
Our object terms are statically typed, so we would
need one of these techniques to interpret dynamically\hyp typed
terms such as those read from a file.

@@

Our main idea is to represent object programs not in an initial algebra
but using the existing coalgebraic structure of the $\lambda$-calculus.
More generally, to squeeze more invariants out of a type system as
simple as Hindley-Milner, we shift the burden of representation and
computation from consumers to producers: encoding object terms as calls
to metalanguage functions (\S\ref{ourapproach}); build dynamic terms
alongside static ones (\S\ref{S:PE-lift}); simulating type functions for
partial evaluation (\S\ref{S:PE-solution}) and CPS
transformation~(\S\ref{S:CPS}).  This shift also underlies fusion,
functionalization, and amortized complexity analysis.

When the metalanguage does provide higher-rank and higher-kind
polymorphism, we can type-check and compile an object term separately
from any interpreters it may be plugged into.

>>>>

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.
It is straightforward to create an algebraic data type, say in OCaml%
%aplas-specific
Fig.~\ref{fig:tag-problem}(a)%
, to
represent object terms such as those in Figure~\ref{fig:object}.
% fig:tag-problem part (a) was here
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
%aplas-specific
\texttt{let test1 = A (L (V VZ), B true)}.
%aplas-else
\begin{comment}
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}
\end{comment}

@@

(b) \begin{code2}
let rec lookup (x::env) = function | VZ   -> x
                                   | VS v -> lookup env v
let rec eval0 env = function
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code2}

(c) \begin{code2}
type u = UB of bool | UA of (u -> u)
\end{code2}

(d) \begin{code2}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code2}
\caption{OCaml code illustrating the tag problem.}
\label{fig:tag-problem}
\end{figure}
% 
Following \cite{WalidICFP02},
we try to implement an interpreter function |eval0|%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(b)%
. It takes
an object term such as |test1| above and gives us its value.
The first argument to |eval0| is the environment, initially empty,
which is the list of values bound to free variables in the
interpreted code.
% fig:tag-problem part (b) was here
If our OCaml-like metalanguage were untyped, the code above would be 
acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \citep[\S1.1--2]{WalidICFP02}.

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(c)%
.
% fig:tag-problem part (c) was here
We can then write a typed interpreter%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(d)
% fig:tag-problem part (d) was here
whose inferred type is |u list -> exp -> u|. Now we can evaluate
%aplas-specific
\texttt{eval [] test1} obtaining |UB true|.
%aplas-else
\begin{comment}
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}
\end{comment}
% fig:tag-problem part (f) was here
The unfortunate tag |UB| in the result reflects that |eval| is a partial
function.  
%aplas-specific
First, the pattern match |with UA f| in the line |A (e1,e2)| is not 
exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term |A (L (V (VS VZ)), B true)|.
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.
%aplas-else
\begin{comment}
First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term
\begin{code}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.
\end{comment}

@@

%aplas-else
\begin{comment}
\subsection{Solutions using fancier types}
\end{comment}

It is commonly thought that to interpret a typed object language in
a typed metalanguage while preserving types is difficult and requires
GADTs or dependent types \citep{taha-tag}.  In fact, this problem
motivated much work on GADTs \cite{xi-guarded,peyton-jones-simple} and
on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
%aplas-else
\begin{comment}
For a metalanguage's type system to allow the well-typed object term
|test1| but disallow the ill-typed object term |test2|, fancier types
such as GADTs or dependent types seem necessary.  
\end{comment}
Yet other type systems
have been proposed to distinguish closed terms like |test1| from open
terms like |test3|
%aplas \cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01},
so that |lookup| never receives an empty environment.  We discuss these
proposals further in \S\ref{related}%
%aplas-specific
.
%aplas-else
\begin{comment}
; here we just note that many
advanced type systems have been devised to ensure statically that an
object term is well-typed and closed.
\end{comment}

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions
\begin{code3}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code3}
which comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
%aplas-specific
\texttt{let testf1 = app (lam varZ) (b true)}
%aplas-else
\begin{comment}
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
\end{comment}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment%
%aplas-specific
, |testf1 ()|, obtaining |true|.
%aplas-else
\begin{comment}
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
\end{comment}
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
%aplas-specific
open term such as \texttt{app (lam (varS varZ)) (b true)}
gives a type- rather than a run-time error.
%aplas-else
\begin{comment}
\begin{code}
open term such as |testf3| below gives a type error rather than
a run-time error
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
\end{comment}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

In sum, by Church\hyp encoding terms using ordinary functions, we
achieve a tagless evaluator for a typed object language in a
metalanguage with a simple 
%aplas
Hindley-Milner 
type system
%aplas \cite{hindley-principal,milner-theory}
.  In this \emph{final} rather
than \emph{initial} approach, both kinds of run-time errors in
\S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter
uses no universal type or pattern matching, it never results in a
run-time error (and is in fact total).  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \cite{WalidICFP02} of error checking.
\Citet{Gluck-jones-optimality} explains deeper technical reasons that
inevitably lead to these performance penalties.

@@

(c) \begin{code2}
testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code2}
\caption{Symantics in Haskell.}
\label{fig:symantics-haskell}
\end{figure}
% fig:symantics-haskell part (a) was here
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
whose inferred type is \texttt{Symantics repr => repr Bool}.
For another example, the classical $\mathit{power}$ function is
%aplas-specific
in Fig.~\ref{fig:symantics-haskell}(b)
% fig:symantics-haskell part (b) was here
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
in Fig.~\ref{fig:symantics-haskell}(c).
% fig:symantics-haskell part (c) was here
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag.
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete implementations.

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation
\begin{code3}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
: \texttt{module EXR = EX(R)}%
.
%% \begin{code}
%% module EXR = EX(R)
%% \end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
% aplas: commented out
\begin{comment}
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
\end{comment}
%aplas: the following line is for aplas only
It is obvious to the compiler that
%
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
%aplas-specific
(The source code shows a total interpreter |L| that measures
the size of each object term).
%
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
These propositions follow immediately from the soundness of the
metalanguage's type system.
% aplas-commented
\begin{comment}
For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the expression
\begin{code}
let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
\begin{code}
app (fix (fun self -> self)) (int 1)
\end{code}

@@

(d) \begin{code2}
.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
   n_4 in self_3) x_1 7>.
\end{code2}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
%
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(b),
% fig:interpreter-C part (b) was here
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
with
%aplas-specific
Fig.~\ref{fig:interpreter-C}(c),
% fig:interpreter-C part (c) was here
gives a code value with many apparent $\beta$-redexes
%aplas-specific
Fig.~\ref{fig:interpreter-C}(d).
% fig:interpreter-C part (d) was here
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and does not
  call the interpreter, unlike the recursive calls to |eval0| and
  |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.
%aplas: the line is for aplas. The original paragraph commented out
The accompanying code shows the corresponding Haskell implemention. 
\begin{comment}
We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
  (TH). The output of our compiler will then be assuredly type-correct TH.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.
\end{comment}

@@

Roughly, a partial evaluator interprets each object term to either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, one might define
the Haskell data type 
%aplas-specific
\texttt{data P0 t = S0 (R t) $\mid$ E0 (C t)}.
We then start to define |P0| as an instance of the |Symantics| class.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.
%aplas-else
\begin{comment}
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

@@

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}
\end{comment}

@@

%aplas-else
\begin{comment}
  A lot of effort has gone into ``typing dynamic typing'': to statically
  type-check dynamically\hyp typed values
  \cite{baars-typing,haskell-list},
  using the host language's type system to varying extents.
  Our object terms are statically typed, so we would
  need one of these techniques to interpret dynamically\hyp typed
  terms such as those read from a file.
\end{comment}

@@

Our main idea is to represent object programs not in an initial algebra
but using the existing coalgebraic structure of the $\lambda$-calculus.
More generally, to squeeze more invariants out of a type system as
simple as Hindley-Milner, we shift the burden of representation and
computation from consumers to producers: encoding object terms as calls
to metalanguage functions (\S\ref{ourapproach}); build dynamic terms
alongside static ones (\S\ref{S:PE-lift}); simulating type functions for
partial evaluation (\S\ref{S:PE-solution}) and CPS
transformation%
%aplas
.
%aplas-else ~(\S\ref{S:CPS}).
This shift also underlies fusion,
functionalization, and amortized complexity analysis.
%aplas-else
\begin{comment}
When the metalanguage does provide higher-rank and higher-kind
polymorphism, we can type-check and compile an object term separately
from any interpreters it may be plugged into.
\end{comment}
